---
title: "Assignment 1"
subtitle: "Data Preparation and Analysis"
author: "Lauren Campbell-Hansen"
format: html
---

Load extra r packages

```{r}
library(tidyverse)
library(haven)

raw_data_dir <- "data/raw/"
processed_data_dir <- "data/processed/202609/"
```

### Assignment

Dataset with one row per person for 2019 all danes age 18-65 who are employed With vars/colomns.

The final dataset must contain:

- `pnr`
- `alder`
- `koen`
- `civ`
- `uddannelse`
- `indkomst`
- `gen_indkomst`
- `forskel_indkomst`
- `uddannelse_alder`

# 1.1 read data in and take a look

Read the BEF, UDDA, IND, HFAUDD, and PRE_SOCIO data sources.

```{r}
bef <-read_dta(paste0(raw_data_dir,"/bef2019.dta")) |> as.data.frame()
udda <-read_dta(paste0(raw_data_dir,"/uddf2024.dta")) |> as.data.frame()
ind <-read_dta(paste0(raw_data_dir,"/ind2019.dta")) |> as.data.frame()


glimpse(bef)
glimpse(udda)
glimpse(ind)

# Bef should contain pnr, alder, koen, civ, uddannelse, indkomst, gen_indkomst, forskel_indkomst and uddannelse_alder
# all contain pnr = merge id
```

## Reading in txt files

```{r}

hfaudd <- read_tsv(
  paste0(raw_data_dir, "/hfaudd_disced15.txt"),
  show_col_types = FALSE
) |>
  as.data.frame()

pre_socio <- read_delim(
  paste0(raw_data_dir, "/pre_socio_klassifikation.txt"),
  delim = ";",
  show_col_types = FALSE
)

pre_socio_c <- pre_socio |>
  select(
    pre_socio = START,
    arbejdsmarked = ANALYSE_KAT
  ) |>
  mutate(
    pre_socio = as.character(pre_socio)
  )

glimpse(hfaudd)
glimpse(pre_socio_c)
```

### 1.2 Categorical variables and variable types

```{r}
# bef = bef_c - just clean version
bef_c <- bef |>
  select(
    pnr,
    alder,
    koen,
    civ,
    foedselsaar = `fødeår`
  ) |>
  filter(between(alder, 18,65)) |>
  mutate(
    pnr = as.character(pnr), #insure all joining variables have compatible types

    koen = recode(
      koen,
      `1` = "Male",
      `2` = "Female"
    ),

    civ = recode(
      civ,
      "U" = "Unmarried",
      "G" = "Married",
      "F" = "Divorced",
      "E" = "Widowed",
      "P" = "Registered partnership",
      "O" = "Dissolved partnership",
      "L" = "Surviving partner"
    )
  )

glimpse (bef_c)

```

I create cleaned versions of the original datasets so that the original objects remain unchanged.

## Prepare person-level education data

The education code must remain available until it has been connected to the education classification table.

```{r}
hfaudd_c <- hfaudd |>
  mutate(
    HFAUDD = str_pad(
      as.character(HFAUDD),
      width = 4,
      side = "left",
      pad = "0"
    )
  )

glimpse(hfaudd_c)

udda_c <- udda |>
  transmute(
    pnr = as.character(pnr),
    
    hfaudd = str_pad(
      as.character(hfaudd),
      width = 4,
      side = "left",
      pad = "0"
    ),
    
    hf_vfra = as.numeric(hf_vfra)
  )

glimpse(udda_c)

```

Attach the human-readable education level to the person-level education data.

```{r}
udda_labelled <- udda_c |>
  left_join(
    hfaudd_c,
    by = c("hfaudd" = "HFAUDD")
  ) |>
  transmute(
    pnr,
    hf_vfra,
    uddannelse = NIVEAU_TEKST
  )

glimpse(udda_labelled)

```

## Prepare income and employment data

```{r}
ind_c <- ind |>
  transmute(
    pnr = as.character(pnr),
    pre_socio = as.character(pre_socio),
    indkomst = as.numeric(perindkialt_13)
  )
pre_socio_c <- pre_socio_c |>
  mutate(
    pre_socio = as.character(pre_socio)
  )

glimpse(ind_c)
glimpse(pre_socio_c)

```

### 1.3 Merging datasets

The person-level datasets are joined using `pnr`. Employment status is attached using the `pre_socio` classification code.

```{r}
dt <- bef_c |>
  left_join(
    udda_labelled,
    by = "pnr"
  ) |>
  left_join(
    ind_c,
    by = "pnr"
  ) |>
  left_join(
    pre_socio_c,
    by = "pre_socio"
  )

glimpse(dt)
```

#### Check employment status

Check which employment categories are present.

```{r}
count(dt, arbejdsmarked)
```

Retain employed individuals only.

```{r}
dt <- dt |>
  filter(arbejdsmarked == "employed")

count(dt, arbejdsmarked)
```

### 1.4 Create new variables

First, missing education information is recoded as `Unknown`.

```{r}
dt <- dt |>
  mutate(
    uddannelse = replace_na(
      uddannelse,
      "Unknown"
    )
  )

count(dt, uddannelse)
```

Calculate age at completion of education.

```{r}
dt <- dt |>
  mutate(
    uddannelse_alder = hf_vfra - foedselsaar
  )

summary(dt$uddannelse_alder)
```

Calculate the mean income for each education group.

```{r}
dt <- dt |>
  group_by(uddannelse) |>
  mutate(
    gen_indkomst = mean(
      indkomst,
      na.rm = TRUE
    )
  ) |>
  ungroup()
```

Calculate the difference between each individual’s income and the mean income for their education group.

```{r}
dt <- dt |>
  mutate(
    forskel_indkomst = indkomst - gen_indkomst
  )
```

### 1.5 Prepare the final dataset

Convert the categorical variables to factors and retain exactly the required columns.

```{r}
dt <- dt |>
  mutate(
    koen = as_factor(koen),
    civ = as_factor(civ),
    uddannelse = as_factor(uddannelse)
  ) |>
  select(
    pnr,
    alder,
    koen,
    civ,
    uddannelse,
    indkomst,
    gen_indkomst,
    forskel_indkomst,
    uddannelse_alder
  )
```

Inspect the completed dataset.

```{r}
glimpse(dt)
summary(dt)
```

Confirm that the dataset contains exactly the required columns.

```{r}
names(dt)
```

### 1.6 Save the final dataset

The finished dataset is saved as a CSV file in the `code` folder.

```{r}
#| label: save-final-dataset

write_csv(
  dt,
  "code/assignment1_campbell_hansen.csv"
)
```

## Codebook for the final dataset

**Population:** Employed Danish individuals aged 18–65 in 2019. The final dataset contains one row per person.

| Column | Type | What it holds | Raw dataset and variable | Data management |
|---------------|---------------|---------------|---------------|---------------|
| `pnr` | character | Unique person identifier used to connect records across datasets | `bef2019.dta`: `pnr`; also present in `uddf2024.dta` and `ind2019.dta` | Converted to character and used as the person-level joining key |
| `alder` | numeric | Age in completed years in 2019 | `bef2019.dta`: `alder` | Only people aged 18–65 were retained |
| `koen` | factor | Sex in human-readable categories | `bef2019.dta`: `koen` | Recoded from `1` and `2` to `Male` and `Female`, then converted to a factor |
| `civ` | factor | Marital status in human-readable categories | `bef2019.dta`: `civ` | Register codes were recoded into readable marital-status categories and converted to a factor |
| `uddannelse` | factor | Highest achieved education level in human-readable words | `uddf2024.dta`: `hfaudd`; `hfaudd_disced15.txt`: `HFAUDD` and `NIVEAU_TEKST` | `hfaudd` was joined to `HFAUDD`; `NIVEAU_TEKST` was retained and missing values were recoded as `Unknown` |
| `indkomst` | numeric | Individual income level | `ind2019.dta`: `perindkialt_13` | Renamed `indkomst` and converted to numeric |
| `gen_indkomst` | numeric | Mean income among employed individuals aged 18–65 with the same education level | Derived from `indkomst` and `uddannelse` | Calculated as the mean income within each education group; missing income values were excluded |
| `forskel_indkomst` | numeric | Difference between individual income and the mean income for the same education group | Derived from `indkomst` and `gen_indkomst` | Calculated as `indkomst - gen_indkomst`; positive values indicate income above the group mean |
| `uddannelse_alder` | numeric | Approximate age at which the education level was achieved | `uddf2024.dta`: `hf_vfra`; `bef2019.dta`: `fødeår` | Calculated as education year minus recorded birth year |
